Introduction to the \proglang{R} package \pkg{TDA}

### 2015-09-15 inst/doc/article.Rnw
version 1.4

\author{Brittany T. Fasy \and Fabrizio Lecci \\ \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Fabrizio Lecci}

\title{Introduction to the \pkg{TDA} package}
\Plaintitle{Introduction to the TDA package}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{Dionysus} C++ library, including a function for 
the persistent homology of the Rips filtration, and for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2013stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which  allows us to identify  the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

@@

@@

<<initial_settings, echo=FALSE>>=
options(stringsAsFactors=FALSE)
options(width=65)
options(prompt="> ")
@

@@

In Section \ref{sec:distances}, we describe how to compute some widely studied functions that, starting from a point cloud, provide some topological information about the underlying space: the distance function (\code{distFct}), the distance to a measure function (\code{dtm}), the k Nearest Neighbor density estimator (\code{knnDE}), the kernel density estimator (\code{kde}), and the kernel distance (\code{kernelDist}).

One approach to TDA is persistent homology \citep{edelsbrunner2010computational}, a method for studying the homology at multiple scales simultaneously. More precisely, it provides a framework and efficient algorithms to quantify the evolution of the topology of a family of nested topological spaces. Given a real-valued function $f$, such as the ones described in Section~\ref{sec:distances}, persistent homology describes how the topology of the lower level sets $\{x: f(x) \leq t \}$ (or superlevel sets $\{x: f(x) \geq t \}$) changes as $t$ increases from $-\infty$ to $\infty$ (or decreases from $\infty$ to $-\infty$). 
This information is encoded in the persistence diagram, a multiset of points in the plane, each corresponding to the birth and death of a homological feature that existed for some interval of $t$.
In Section \ref{sec:persistent}, we describe how to compute persistence diagrams of sublevel sets (or superlevel sets) of functions evaluated over a grid of points, using \code{gridDiag}, and how to compute the persistence diagram of the Rips filtration built on top of a point cloud, using \code{ripsDiag}. Our implementation relies on \pkg{Dionysus}, a \proglang{C++} library developed by Dmitriy Morozov and freely available at http://mrzv.org/software/dionysus/ .

One of the key challenges in persistent homology is to find a way to isolate the points of the persistence diagram representing the topological noise. In \cite{fasy2014statistical}, \cite{chazal2013stochastic}, and \cite{chazal2014robust}, we propose several statistical methods to construct confidence sets for persistence diagrams and other summary functions that allow us to separate topological signal from topological noise.
Some of these methods are implemented and described in Section \ref{sec:persistent}.

Finally, the \pkg{TDA} package provides the implementation of an algorithm for density clustering. 
This method allows us to identify and visualize the spatial organization of the data, without specific knowledge about the data generating mechanism and in particular without any a priori information about the number of clusters. 
In Section \ref{sec:density}, we describe the function \code{clusterTree}, that, given a density estimator, encodes the hierarchy of the connected components of its superlevel sets into a dendrogram, the cluster tree. \citep{KpoLux11, kent2013}.

@@

As a first toy example to using the \pkg{TDA} package, we show how to compute
distance functions and density estimators over a grid of points.
The setting is the typical one in TDA: a set of points $X=\{x_1, \dots, x_n\}$ 
has been sampled from some distribution $P$ supported on a manifold and we 
are interested in recovering the topological features of the underlying manifold
by studying some functions of the data. 
The following code generates a sample of 400 points from the unit circle and constructs
a grid of points over which we will evaluate the functions.

<<eq1, echo=TRUE>>=
library("TDA")
X = circleUnif(400)

Xlim=c(-1.6, 1.6);   Ylim=c(-1.7, 1.7);   by=0.065

Xseq=seq(Xlim[1], Xlim[2], by=by)
Yseq=seq(Ylim[1], Ylim[2], by=by)
Grid=expand.grid(Xseq,Yseq)
@

The \pkg{TDA} package provides implementations of the following functions:
\begin{itemize}
\item The distance function is defined for each $y \in \mathbb{R}^d$ as
$
\Delta(y)= \inf_{x \in X} \| x-y \|_2
$
and is computed for each point of the \code{Grid} with the following code:
<<eq2, echo=TRUE>>=
distance= distFct(X=X, Grid=Grid)
@
\item Given a probability measure $P$, the distance to measure (DTM) is defined for each $y \in \mathbb{R}^d$ as
$$
d_{m_0}(y) =\sqrt{ \frac{1}{m_0}\int_0^{m_0} ( G_y^{-1}(u))^2 du},
$$
where
$G_y(t) = P( \Vert X-y \Vert \leq t)$ and $0<m_0<1$ is a smoothing parameter. For more details
see \cite{chazal2011geometric}.\\
Given $X=\{x_1, \dots, x_n\}$, the empirical version of the DTM is
$$
\hat d_{m_0}(y) = \sqrt{ \frac{1}{k} \sum_{x_i \in N_k(y)} \Vert x_i-y \Vert^2 },
$$
where $k= \lceil m_0 n \rceil$ and $N_k(y)$ is the 
set containing the $k$ nearest neighbors 
of $y$ among $x_1, \ldots, x_n$.
The DTM is computed for each point of the \code{Grid} with the following code:
<<eq3, echo=TRUE>>=
m0=0.1
DTM=dtm(X=X, Grid=Grid, m0=m0)
@
\item The $k$ Nearest Neighbor density estimator, for each $y \in \mathbb{R}^d$, is defined by
$$
\hat\delta_k(y)=\frac{k}{n \; v_d \; r_k^d(y)},
$$
where $v_n$ is the volume of the Euclidean $d$ dimensional unit ball and $r_k^d(x)$ is the Euclidean distance form point $x$ to its $k$th closest neighbor among the points of $X$.
It is computed for each point of the \code{Grid} with the following code:
<<eq4, echo=TRUE>>=
k=60
kNN=knnDE(X=X, Grid=Grid, k=k)
@
\item The Gaussian Kernel Density Estimator (KDE), for each $y \in \mathbb{R}^d$, is defined by
$$
\hat p_h(y)=\frac{1}{n (\sqrt{2 \pi} h )^d} \sum_{i=1}^n \exp \left( \frac{- \Vert y-x_i \Vert_2^2}{2h^2} \right).
$$
where $h$ is a smoothing parameter.
It is computed for each point of the \code{Grid} with the following code:
<<eq5, echo=TRUE>>=
h=0.3
KDE= kde(X=X, Grid=Grid, h=h)
@
\item The Kernel distance estimator, for each $y \in \mathbb{R}^d$, is defined by
$$
\hat \kappa_h(y)=\sqrt{ \frac{1}{n^2} \sum_{i=1}^n\sum_{j=1}^n K_h(x_i, x_j) + K_h(y,y) - 2 \frac{1}{n} \sum_{i=1}^n K_h(y,x_i)  },
$$
where $K_h(x,y)=\exp\left( \frac{- \Vert x-y \Vert_2^2}{2h^2} \right)$ is the Gaussian Kernel with smoothing parameter $h$.
The Kernel distance is computed for each point of the \code{Grid} with the following code:
<<eq6, echo=TRUE>>=
h=0.3
Kdist= kernelDist(X=X, Grid=Grid, h=h)
@
\end{itemize}
For this 2 dimensional example, we can visualize the functions using \code{persp} form the \pkg{graphics} package. For example the following code produces the KDE plot in Figure \ref{fig:eq8}:

<<eq7, echo=TRUE, eval=FALSE, height=6, width=6>>=
persp(Xseq,Yseq,matrix(KDE,ncol=length(Yseq), nrow=length(Xseq)),
xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, 
main="KDE", d=0.5, scale=FALSE, expand=3, shade=0.9)
@

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=6in, height=4.6in}
\begin{center}
<<eq8, echo=FALSE, eval=TRUE, fig=TRUE, height=5, width=6.5>>=
par(mfrow=c(2,3))
par(mai=c(0.6,0.25,0.3,0.1))
	plot(X, pch=16, cex=0.6, xlab="", ylab="", main="Sample X")

par(mai=c(0.1,0.15,0.1,0.1))
	persp(Xseq,Yseq,matrix(distance,ncol=length(Yseq), nrow=length(Xseq)), xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, main="Distance Function", d=0.5, scale=FALSE, expand=1, shade=0.9)

	persp(Xseq,Yseq,matrix(DTM,ncol=length(Yseq), nrow=length(Xseq)), xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, main="DTM", d=0.5, scale=FALSE, expand=1, shade=0.9)

	persp(Xseq,Yseq,matrix(kNN,ncol=length(Yseq), nrow=length(Xseq)), xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, main="kNN", d=0.5, scale=FALSE, expand=3, shade=0.9)

	persp(Xseq,Yseq,matrix(KDE,ncol=length(Yseq), nrow=length(Xseq)), xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, main="KDE", d=0.5, scale=FALSE, expand=3, shade=0.9)

	persp(Xseq,Yseq,matrix(Kdist,ncol=length(Yseq), nrow=length(Xseq)), xlab="",ylab="",zlab="",theta=-20,phi=35, ltheta=50, col=2, border=NA, main="Kernel Distance", d=0.5, scale=FALSE, expand=5, shade=0.9)
@
\end{center}
\vspace{-1cm}
\caption{distance functions and density estimators evaluated over a grid of points.}
\label{fig:eq8}
\end{figure}

@@

We can construct a $(1-\alpha)$ confidence band for a function using the bootstrap algorithm, which we briefly describe using the kernel density estimator:
\begin{enumerate}
\item Given a sample $X=\{x_1, \dots, x_n\}$, compute the kernel density estimator $\hat p_h$;
\item Draw $X^*=\{x_1^*, \dots, x_n^*\}$ from $X=\{x_1, \dots, x_n \}$ (with replacement), and compute $\theta^*=\sqrt{n} \Vert \hat p_h^*(x) - \hat p_h(x) \Vert_\infty$, where $\hat p_h^*$ is the density estimator computed using $X^*$;
\item Repeat the previous step $B$ times to obtain $\theta_{1}^*, \dots, \theta_{B}^*$;
\item Compute 
$
q_{\alpha} = \inf \left \{q: \frac{1}{B} \sum_{j=1}^B I(\theta_j^* \geq q) \leq \alpha \right\};
$
\item The $(1-\alpha)$ confidence band for $\mathbb{E}[\hat p_h]$ is
$
\left[\hat p_h - \frac{q_{\alpha}}{\sqrt{n}} \,, \, \hat p_h+ \frac{q_{\alpha}}{\sqrt{n}}\right] .
$
\end{enumerate}

@@

<<eq9, echo=TRUE, eval=TRUE>>=
band=bootstrapBand(X=X, FUN=kde, Grid=Grid, B=100, 
                   parallel=F, alpha=0.1, h=h)
@

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=5in, height=2.5in}
\begin{center}
<<eq10, echo=FALSE, eval=TRUE, fig=TRUE, width=8, height=4>>=
posYgrid=which(Grid[,2]>0)
posYseq=which(Yseq>0)

par(mfrow=c(1,2))
par(mai=c(0.5,0.25,0.3,0.1))

persp(Xseq,Yseq,matrix(band$band[,1],ncol=length(Yseq), nrow=length(Xseq)),zlim=c(0,max(band$band[posYgrid,2])), ylim = range(Yseq),
xlab="",ylab="",zlab="",theta=0,phi=25, ltheta=50, col="pink", border=NA, d=0.5, scale=FALSE, expand=3, shade=0.9,box=F)

persp(Xseq,Yseq[posYseq],matrix(band$band[posYgrid,1],ncol=length(Yseq[posYseq]), nrow=length(Xseq)),zlim=c(0,max(band$band[posYgrid,2])), ylim = c(-0.5,1.7),
xlab="",ylab="",zlab="",theta=0,phi=25, ltheta=50, col="pink", border=NA, d=0.5, scale=FALSE, expand=3, shade=0.9,box=F)

@@

persp(Xseq,Yseq[posYseq],matrix(band$fun[posYgrid],ncol=length(Yseq[posYseq]), nrow=length(Xseq)),zlim=c(0,max(band$band[posYgrid,2])), ylim = c(-0.5,1.7),,
xlab="",ylab="",zlab="",theta=0,phi=25, ltheta=50, col="red", border=NA, d=0.5, scale=FALSE, expand=3, shade=0.9,box=F)

@@

persp(Xseq,Yseq[posYseq],matrix(band$band[posYgrid,2],ncol=length(Yseq[posYseq]), nrow=length(Xseq)), zlim=c(0,max(band$band[posYgrid,2])), ylim = c(-0.5,1.7),
xlab="",ylab="",zlab="",theta=0,phi=25, ltheta=50, col="pink", border=NA, 
main="", d=0.5, scale=FALSE, expand=3, shade=0.9, box=F)
@
\end{center}
\vspace{-1cm}
\caption{the 90\% confidence band for $\mathbb{E}[\hat p_h]$ has the form $[\ell, u]= \left[\hat p_h - q_{\alpha}/\sqrt{n} \,, \, \hat p_h+ q_{\alpha}/\sqrt{n}\right]$. The plot on the right shows a section of the functions: the red surface is the KDE $\hat p_h$; the pink surfaces are $\ell$ and $u$.}
\label{fig:eq10}
\end{figure}
% PLOT END

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. \code{gridDiag} currently supports grids in dimension 1,2, and 3. Future releases of the \pkg{TDA} package will extend its functionalities.

The following code computes the persistent homology of the superlevel sets (\code{sublevel=FALSE}) of the kernel density estimator (\code{FUN=kde}, \code{h=0.3}) using the point cloud stored in the matrix~\code{X} from the previous example.
The same code would work for the other functions defined in Section \ref{sec:distances} (it is sufficient to replace \code{kde} and its smoothing parameter \code{h} with another function and its parameters). 
The function \code{gridDiag} returns an object of the class \code{"diagram"}.  
The other inputs are the features of the grid over which the \code{kde} is evaluated (\code{Xlim}, \code{Ylim} and \code{by}), the smoothing parameter \code{h}, and a logical variable that indicates whether a progress bar should be printed (\code{printStatus}).

<<eq11, echo=TRUE, eval=TRUE>>=
Diag = gridDiag( X=X, FUN=kde, h=0.3, Xlim=Xlim, Ylim=Ylim, by=by, 
                 sublevel=F, printStatus=F ) 
@

We plot the data and the the diagram, using the function \code{plot}, implemented as a standard \code{S3} method for objects of the class \code{"diagram"}. The following command produces the third plot in Figure \ref{fig:eq11b}.

<<eq11b, echo=TRUE, eval=FALSE>>=
plot(Diag, band=2*band$width, main="KDE Diagram")
@

>>>>

\author{Brittany T. Fasy, Jisu Kim, Fabrizio Lecci \and Cl\'ement Maria \\ \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria}

\title{Introduction to the \proglang{R} package \pkg{TDA}}
\Plaintitle{Introduction to the R package TDA}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{C++} libraries \pkg{GUDHI}, \pkg{Dionysus} and \pkg{PHAT}, 
including a function for the persistent homology of the Rips filtration, 
and one for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2014stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which  allows us to identify  the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

@@

  Jisu Kim\\
  Department of Statistics\\
  Carnegie Mellon University\\
  Email: \email{jisuk1@andrew.cmu.edu}\\

  Cl\'ement Maria\\
  Geometrica Group\\
  INRIA Sophia Antipolis-M\'editerran\'ee\\
  Email: \email{clement.maria@inria.fr}\\

@@

<<initial_settings, echo=FALSE>>=
options(stringsAsFactors = FALSE)
options(width = 65)
options(prompt = "> ")
@

@@

One approach to TDA is persistent homology \citep{edelsbrunner2010computational}, a method for studying the homology at multiple scales simultaneously. More precisely, it provides a framework and efficient algorithms to quantify the evolution of the topology of a family of nested topological spaces. Given a real-valued function $f$, such as the ones described in Section~\ref{sec:distances}, persistent homology describes how the topology of the lower level sets $\{x: f(x) \leq t \}$ (or superlevel sets $\{x: f(x) \geq t \}$) changes as $t$ increases from $-\infty$ to $\infty$ (or decreases from $\infty$ to $-\infty$). 
This information is encoded in the persistence diagram, a multiset of points in the plane, each corresponding to the birth and death of a homological feature that existed for some interval of $t$.

%There is a growing list of software available to compute persistent homology.
%\textsf{\textbf{javaPlex}}\footnote{https://code.google.com/p/javaplex/} is a
%Java software package developed by the Computational Topology workgroup at
%Stanford University;
%\textsf{\textbf{phom}}\footnote{
%http://cran.r-project.org/web/packages/phom/index.html} is an R package written
%by Andrew Tausz;
%\textsf{\textbf{PHAT}}\footnote{https://code.google.com/p/phat/} is a C++
%library developed at the Institute of Science and Technology Austria;
%\textsf{\textbf{Dionysus}}\footnote{http://www.mrzv.org/software/dionysus/} is
%a
%C++ library written and maintained by Dmitriy Morozov;
%finally, \textsf{\textbf{GUDHI}}\footnote{https://project.inria.fr/gudhi/} is
%new born project hosted by INRIA and whose goal is the development and
%implementation of new algorithms for geometric understanding in high
%dimensions.
%Preliminary results show that \textsf{\textbf{GUDHI}} outperforms its major
%competitors \textsf{\textbf{PHAT}} and \textsf{\textbf{Dionysus}}; see
%\cite{boissonnat2013compressed}.

This paper is devoted to the presentation of the \proglang{R} package \pkg{TDA},
which provides a user-friendly interface for the efficient algorithms of the 
\proglang{C++} libraries \pkg{GUDHI} \citep{gudhi_stpcoh}, \pkg{Dionysus} \citep{dionysus_C}, and \pkg{PHAT} \citep{phat_C}.

In Section \ref{sec:distances}, we describe how to compute some widely studied functions that, starting from a point cloud, provide some topological information about the underlying space: the distance function (\code{distFct}), the distance to a measure function (\code{dtm}), the k Nearest Neighbor density estimator (\code{knnDE}), the kernel density estimator (\code{kde}), and the kernel distance (\code{kernelDist}).
Section \ref{sec:persistent} is devoted to the computation of persistence diagrams: the function \code{gridDiag} can be used to compute persistent homology of sublevel sets (or superlevel sets) of functions evaluated over a grid of points; the function \code{ripsDiag} returns the persistence diagram of the Rips filtration built on top of a point cloud.

One of the key challenges in persistent homology is to find a way to isolate the points of the persistence diagram representing the topological noise.
Statistical methods for persistent homology provide an alternative to its exact
computation. Knowing with high confidence that an approximated persistence
diagrams is close to the true--computationally infeasible--diagram is often
enough for practical purposes. 
\cite{fasy2014statistical}, \cite{chazal2014stochastic}, and \cite{chazal2014robust} propose several statistical methods to construct confidence sets for persistence diagrams and other summary functions that allow us to separate topological signal from topological noise.
The methods are implemented in the \pkg{TDA} package and described in Section \ref{sec:persistent}.

Finally, the \pkg{TDA} package provides the implementation of an algorithm for density clustering. 
This method allows us to identify and visualize the spatial organization of the data, without specific knowledge about the data generating mechanism and in particular without any a priori information about the number of clusters. 
In Section \ref{sec:density}, we describe the function \code{clusterTree}, that, given a density estimator, encodes the hierarchy of the connected components of its superlevel sets into a dendrogram, the cluster tree \citep{KpoLux11, kent2013}.

@@

As a first toy example to using the \pkg{TDA} package, we show how to compute
distance functions and density estimators over a grid of points.
The setting is the typical one in TDA: a set of points $X=\{x_1, \dots, x_n\} \subset \mathbb{R}^d$ 
has been sampled from some distribution $P$ and we 
are interested in recovering the topological features of the underlying space
by studying some functions of the data. 
The following code generates a sample of 400 points from the unit circle and constructs
a grid of points over which we will evaluate the functions.

<<eq1, echo=TRUE>>=
library("TDA")
X <- circleUnif(400)

Xlim <- c(-1.6, 1.6);  Ylim <- c(-1.7, 1.7);  by <- 0.065

Xseq <- seq(Xlim[1], Xlim[2], by = by)
Yseq <- seq(Ylim[1], Ylim[2], by = by)
Grid <- expand.grid(Xseq, Yseq)
@

The \pkg{TDA} package provides implementations of the following functions:

\begin{itemize}
\item The distance function is defined for each $y \in \mathbb{R}^d$ as
$
\Delta(y)= \inf_{x \in X} \| x-y \|_2
$
and is computed for each point of the \code{Grid} with the following code:

<<eq2, echo=TRUE>>=
distance <- distFct(X = X, Grid = Grid)
@
\item Given a probability measure $P$, the distance to measure (DTM) is defined for each $y \in \mathbb{R}^d$ as
$$
d_{m_0}(y) =\sqrt{ \frac{1}{m_0}\int_0^{m_0} ( G_y^{-1}(u))^2 du},
$$
where
$G_y(t) = P( \Vert X-y \Vert \leq t)$ and $0<m_0<1$ is a smoothing parameter. 
The DTM can be seen as a smoothed version of the distance function.
For more details
see \cite{chazal2011geometric}.\\
Given $X=\{x_1, \dots, x_n\}$, the empirical version of the DTM is
$$
\hat d_{m_0}(y) = \sqrt{ \frac{1}{k} \sum_{x_i \in N_k(y)} \Vert x_i-y \Vert^2 },
$$
where $k= \lceil m_0 n \rceil$ and $N_k(y)$ is the 
set containing the $k$ nearest neighbors 
of $y$ among $x_1, \ldots, x_n$.
The DTM is computed for each point of the \code{Grid} with the following code:

<<eq3, echo=TRUE>>=
m0 <- 0.1
DTM <- dtm(X = X, Grid = Grid, m0 = m0)
@
\item The $k$ Nearest Neighbor density estimator, for each $y \in \mathbb{R}^d$, is defined as
$$
\hat\delta_k(y)=\frac{k}{n \; v_d \; r_k^d(y)},
$$
where $v_n$ is the volume of the Euclidean $d$ dimensional unit ball and $r_k^d(x)$ is the Euclidean distance form point $x$ to its $k$th closest neighbor among the points of $X$.
It is computed for each point of the \code{Grid} with the following code:

<<eq4, echo=TRUE>>=
k <- 60
kNN <- knnDE(X = X, Grid = Grid, k = k)
@
\item The Gaussian Kernel Density Estimator (KDE), for each $y \in \mathbb{R}^d$, is defined as
$$
\hat p_h(y)=\frac{1}{n (\sqrt{2 \pi} h )^d} \sum_{i=1}^n \exp \left( \frac{- \Vert y-x_i \Vert_2^2}{2h^2} \right).
$$
where $h$ is a smoothing parameter.
It is computed for each point of the \code{Grid} with the following code:

<<eq5, echo=TRUE>>=
h <- 0.3
KDE <- kde(X = X, Grid = Grid, h = h)
@
\item The Kernel distance estimator, for each $y \in \mathbb{R}^d$, is defined as
$$
\hat \kappa_h(y)=\sqrt{ \frac{1}{n^2} \sum_{i=1}^n\sum_{j=1}^n K_h(x_i, x_j) + K_h(y,y) - 2 \frac{1}{n} \sum_{i=1}^n K_h(y,x_i)  },
$$
where $K_h(x,y)=\exp\left( \frac{- \Vert x-y \Vert_2^2}{2h^2} \right)$ is the Gaussian Kernel with smoothing parameter $h$.
The Kernel distance is computed for each point of the \code{Grid} with the following code:

<<eq6, echo=TRUE>>=
h <- 0.3
Kdist <- kernelDist(X = X, Grid = Grid, h = h)
@
\end{itemize}
For this 2 dimensional example, we can visualize the functions using \code{persp} form the \pkg{graphics} package. For example the following code produces the KDE plot in Figure \ref{fig:eq8}:

<<eq7, echo=TRUE, eval=FALSE, height=6, width=6>>=
persp(Xseq, Yseq,
      matrix(KDE, ncol = length(Yseq), nrow = length(Xseq)), xlab = "",
      ylab = "", zlab = "", theta = -20, phi = 35, ltheta = 50,
      col = 2, border = NA, main = "KDE", d = 0.5, scale = FALSE,
      expand = 3, shade = 0.9)
@

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=6in, height=4.6in}
\begin{center}
<<eq8, echo=FALSE, eval=TRUE, fig=TRUE, height=5, width=6.5>>=
par(mfrow = c(2, 3))
par(mai = c(0.6, 0.25, 0.3, 0.1))
plot(X, pch = 16, cex = 0.6, xlab = "", ylab = "", main = "Sample X")

par(mai = c(0.1, 0.15, 0.1, 0.1))
persp(Xseq, Yseq,
      matrix(distance, ncol = length(Yseq), nrow = length(Xseq)),
      xlab = "", ylab = "", zlab = "", theta = -20, phi = 35,
      ltheta = 50, col = 2, border = NA, main = "Distance Function",
      d = 0.5, scale = FALSE, expand = 1, shade = 0.9)
persp(Xseq, Yseq,
      matrix(DTM, ncol = length(Yseq), nrow = length(Xseq)), xlab = "",
      ylab = "", zlab = "", theta = -20, phi = 35, ltheta = 50,
      col = 2, border = NA, main = "DTM", d = 0.5, scale = FALSE,
      expand = 1, shade = 0.9)
persp(Xseq, Yseq,
      matrix(kNN, ncol = length(Yseq), nrow = length(Xseq)), xlab = "",
      ylab = "", zlab = "", theta = -20, phi = 35, ltheta = 50,
      col = 2, border = NA, main = "kNN", d = 0.5, scale = FALSE,
      expand = 3, shade = 0.9)
persp(Xseq, Yseq,
      matrix(KDE, ncol = length(Yseq), nrow = length(Xseq)), xlab = "",
      ylab = "", zlab = "", theta = -20, phi = 35, ltheta = 50,
      col = 2, border = NA, main = "KDE", d = 0.5, scale = FALSE,
      expand = 3, shade = 0.9)
persp(Xseq, Yseq,
      matrix(Kdist, ncol = length(Yseq), nrow = length(Xseq)),
      xlab = "", ylab = "", zlab = "", theta = -20, phi = 35,
      ltheta = 50, col = 2, border = NA, main = "Kernel Distance",
      d = 0.5, scale = FALSE, expand = 5, shade = 0.9)
@
\end{center}
\vspace{-1cm}
\caption{distance functions and density estimators evaluated over a grid of points.}
\label{fig:eq8}
\end{figure}

@@

We can construct a $(1-\alpha)$ confidence band for a function using the bootstrap algorithm, which we briefly describe using the kernel density estimator:

\vspace{0.2cm}

\begin{enumerate}
\item Given a sample $X=\{x_1, \dots, x_n\}$, compute the kernel density estimator $\hat p_h$;
\item Draw $X^*=\{x_1^*, \dots, x_n^*\}$ from $X=\{x_1, \dots, x_n \}$ (with replacement), and compute $\theta^*=\sqrt{n} \Vert \hat p_h^*(x) - \hat p_h(x) \Vert_\infty$, where $\hat p_h^*$ is the density estimator computed using $X^*$;
\item Repeat the previous step $B$ times to obtain $\theta_{1}^*, \dots, \theta_{B}^*$;
\item Compute 
$
q_{\alpha} = \inf \left \{q: \frac{1}{B} \sum_{j=1}^B I(\theta_j^* \geq q) \leq \alpha \right\};
$
\item The $(1-\alpha)$ confidence band for $\mathbb{E}[\hat p_h]$ is
$
\left[\hat p_h - \frac{q_{\alpha}}{\sqrt{n}} \,, \, \hat p_h+ \frac{q_{\alpha}}{\sqrt{n}}\right] .
$
\end{enumerate}

\vspace{0.2cm}

@@

<<eq9, echo=TRUE, eval=TRUE>>=
band <- bootstrapBand(X = X, FUN = kde, Grid = Grid, B = 100,
                      parallel = FALSE, alpha = 0.1, h = h)
@

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=5in, height=2.5in}
\begin{center}
<<eq10, echo=FALSE, eval=TRUE, fig=TRUE, width=8, height=4>>=
posYgrid <- which(Grid[, 2] > 0)
posYseq <- which(Yseq > 0)

par(mfrow = c(1, 2))
par(mai = c(0.5, 0.25, 0.3, 0.1))

persp(Xseq, Yseq, matrix(band[["band"]][, 1], ncol = length(Yseq),
      nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = range(Yseq), xlab = "", ylab = "", zlab="",theta = 0,
      phi = 25, ltheta = 50, col = "pink", border = NA, d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9, box = FALSE)

persp(Xseq, Yseq[posYseq], matrix(band[["band"]][posYgrid, 1],
      ncol = length(Yseq[posYseq]), nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = c(-0.5, 1.7), xlab = "", ylab = "", zlab = "", theta = 0,
      phi = 25, ltheta = 50, col = "pink", border = NA, d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9, box = FALSE)

@@

persp(Xseq, Yseq[posYseq], matrix(band[["fun"]][posYgrid],
      ncol = length(Yseq[posYseq]), nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = c(-0.5, 1.7), xlab = "", ylab = "", zlab = "", theta = 0,
      phi = 25, ltheta = 50, col = "red", border = NA, d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9, box = FALSE)

@@

persp(Xseq, Yseq[posYseq], matrix(band[["band"]][posYgrid, 2],
      ncol = length(Yseq[posYseq]), nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = c(-0.5, 1.7), xlab = "", ylab = "", zlab = "", theta = 0,
      phi = 25, ltheta = 50, col = "pink", border = NA, main = "",
      d = 0.5, scale = FALSE, expand = 3, shade = 0.9, box = FALSE)
@
\end{center}
\vspace{-1cm}
\caption{the 90\% confidence band for $\mathbb{E}[\hat p_h]$ has the form $[\ell, u]= \left[\hat p_h - q_{\alpha}/\sqrt{n} \,, \, \hat p_h+ q_{\alpha}/\sqrt{n}\right]$. The plot on the right shows a section of the functions: the red surface is the KDE $\hat p_h$; the pink surfaces are $\ell$ and $u$.}
\label{fig:eq10}
\end{figure}
% PLOT END

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. From version 1.2, \code{gridDiag} works in arbitrary dimension.
The core of the function is written in \proglang{C++} and the user can choose to compute persistence diagrams using either the \pkg{Dionysus} library or the \pkg{PHAT} library.

The following code computes the persistent homology of the superlevel sets \linebreak (\code{sublevel=FALSE}) of the kernel density estimator (\code{FUN=kde}, \code{h=0.3}) using the point cloud stored in the matrix~\code{X} from the previous example.
The same code would work for the other functions defined in Section \ref{sec:distances} (it is sufficient to replace \code{kde} and its smoothing parameter \code{h} with another function and the corresponding parameter). 
The function \code{gridDiag} returns an object of the class \code{"diagram"}.  
The other inputs are the features of the grid over which the \code{kde} is evaluated (\code{lim} and \code{by}), the smoothing parameter \code{h}, and a logical variable that indicates whether a progress bar should be printed (\code{printProgress}).

<<eq11, echo=TRUE, eval=TRUE>>=
Diag <- gridDiag(X = X, FUN = kde, h = 0.3, lim = cbind(Xlim, Ylim),
                 by = by, sublevel = FALSE, library = "Dionysus",
                 printProgress = FALSE)
@

We plot the data and the diagram, using the function \code{plot}, implemented as a standard \code{S3} method for objects of the class \code{"diagram"}. The following command produces the third plot in Figure \ref{fig:eq11b}.

<<eq11b, echo=TRUE, eval=FALSE>>=
plot(Diag[["diagram"]], band = 2 * band[["width"]],
     main = "KDE Diagram")
@
